### 2010-09-21 Documentation/part2.tex
*** empty log message ***
Improved documenation of #inverse_beta
Changes in manual for version 3.1.7

\begin{table}[t]
\begin{center}
\begin{tabular}{|cllc|}
\hline
{\tt @builtin} & Description & Parameters & Min. version \\
\hline
1 & Neutrino factory $\mu^+$ decay & {\tt @parent\_energy} [GeV], & 2.0 \\
& & {\tt @stored\_muons} &  \\
2 & Neutrino factory $\mu^-$ decay & {\tt @parent\_energy} [GeV], & 2.0 \\
& & {\tt @stored\_muons} &  \\
3 & Beta beam, inv. beta decay & {\tt @end\_point} [GeV], & 3.0 \\
& & {\tt @stored\_ions}, &  \\
& & {\tt @gamma} & \\
4 & Beta beam, beta decay & {\tt @end\_point} [GeV], & 3.0 \\
& & {\tt @stored\_ions}, &  \\
& & {\tt @gamma} & \\
\hline
\end{tabular}
\end{center}
\mycaption{\label{tab:binfluxes} Built-in fluxes currently supported by \GLOBES . For details on
the beta beams, see \Ref~\cite{Huber:2005jk}. }
\end{table}

@@

\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}\index{norm}{Flux!file}
The software assumes that the given flux file has seven columns and
501 lines with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by \GLOBES. In general, it is advisable to provide the flux
between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}),
since this is the energy range considered in the simulation. However,
if part of this interval is omitted in the flux file, zero will be
used there. If some neutrino flavors are not used in the simulation,
the corresponding columns in the flux file have to be filled nevertheless,
\eg\ with zeros.

@@

Cross sections in \GLOBES\ are given as total cross section divided by energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines of the form
\index{norm}{Cross section!file}
\index{aedl}{cross@{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Cross sections for unused neutrino flavors have to be filled with zeros, and
can not just be omitted.

@@

\item[Bin level] This level is determined by the experiment and its analysis. 
Note that energy
bin sizes much smaller than the energy resolution will not improve the results. The energy bin range and the number of energy bins always have to be specified. For the case of large values of the integrand in \eq~(\ref{eq:simple_int}) at the energy range limits, it is recommended to exceed the analysis energy window by about three times the energy calibration error in order 
to avoid cutoff effects.

@@

The key idea is to use a ``flat'' model, \ie\ the integrand 
of~\eq~\ref{eq:simple_int}  is well approximated by being piecewise
constant in each sampling step. This is a good approximation as long as
\begin{itemize}
\item
 No details are lost, \ie\ the spacing of sampling points 
is smaller than the energy resolution.
\item
The edges are treated correctly.
\item
 The neutrino oscillations are slow on a scale of the sampling point distance.
\end{itemize}
In this case, \eq~(\ref{eq:simple_int}) is reduced to
\begin{equation}
\label{eq:algo_one}
n_i^c=N/L^2 \, \sum_{j=1}^N \,  \Phi^c(E_j)\,
P^c(E_j)\,
\sigma^c(E_j)\,
K_i^c(E_j) \, \Delta E_j \,.
\end{equation}
The advantages of this algorithm are obvious: All factors
independent of the oscillation parameters have to be only evaluated once
at values of $E$ which are known in advance, which means that they can be put 
into a look-up table. In addition, the probability
has to be only evaluated at previously known values of the energy, which
makes it possible to compute the transition amplitudes for all channels
simultaneously. One assumption is that all involved factors are piece-wise
constant, \ie, they hardly change within each bin. This 
assumption seems to be very restrictive, which is however not quite correct.
First of all, if one analyzes simulated data (which are simulated with the same algorithm), the errors
will cancel between the simulated and fitted data. Second, and more important,
this algorithm is just a very basic integration routine\footnote{It is planned 
for the future to implement something like a Gau\ss-Kronrod scheme as an
alternative here.} and converges to the true result for decreasing step size.
Thus if the number of sampling points is large enough this algorithm is very
accurate. Bin-based energy smearing is selected by
%
\index{aedl}{energy@{\tt energy}!type@{\tt "@type}}
%
\begin{quote}
{\tt \tb @type = 1}
\end{quote}
within the {\tt \#energy} environment.
The computation of the bin kernel $K_i^c$ is performed
by \GLOBES. Thus, it requires that the number of bins
{\tt \$bins} and the minimum energy {\tt \$emin} and maximum energy {\tt \$emax} are given in case of equidistant bins.
%
\index{norm}{Energy!resolution}
\index{norm}{Energy!resolution function}
%
As far as the parameterization for the energy resolution function 
$R^c(E,E')$ in \eq~(\ref{eq:kernel}) is concerned, the algorithm uses
a Gau\ss ian
\begin{equation}
R^c(E,E')=\frac{1}{\sigma(E)\,\sqrt{2\pi}}\,e^{-\frac{(E-E')^2}{2\sigma^2(E)}} \, .
\end{equation} 
%
\index{aedl}{energy@{\tt energy}!sigma function@{\tt "@sigma\_function}}
%
There are several energy resolution functions available, where by default
{\tt \#standard} is used:
\begin{quote}
{\tt \tb @sigma\_function = \#standard} 
\end{quote}
%
\index{aedl}{energy@{\tt energy}!standard@{\tt \#standard}}
%
The energy resolution function {\tt \#standard} is defined by
\begin{equation}
\label{eq:sigma_e}
\sigma(E)=\alpha\cdot E + \beta \cdot \sqrt{E} +\gamma\, ,
\end{equation}
where the parameters $\alpha, \beta$ and $\gamma$ are provided by the user:
\begin{quote}
{\tt \tb @sigma\_e = \{0.15, 0.0, 0.0\}}
\end{quote}
Currently, another possible choice for {\tt @sigma\_function} is {\tt \#inverse\_beta},
%
\index{aedl}{energy@{\tt energy}!inverse beta@{\tt \#inverse\_beta}}
%
which only uses the parameter $\alpha$. It is defined by
\begin{equation}
\sigma(E)= \left\{\begin{array}{cl}
 \alpha \cdot \sqrt{1000}^{-1}\,\sqrt{x-8\cdot10^{-4}}\,,&\mathrm{for}\,\, 
x>1.8\cdot10^{-3}\\
\alpha\cdot10^{-3} \,,&\mathrm{for}\,\, x \leq 1.8\cdot10^{-3}
\end{array} \right.
\end{equation}
The somewhat complicated form is due to the fact that inverse $\beta$-decay
has a neutrino threshold of $1.8\,\mathrm{MeV}$ and that a neutrino
at threshold already produces $\simeq 1\,\mathrm{MeV}$ visible energy in
the detector (for more details see \eg~\cite{Huber:2003pm}).

@@

In order to ensure that very fast oscillations do not lead to 
aliasing,\index{norm}{Aliasing}
it is possible to impose a low-pass filter already during the calculation
of the probabilities itself. This highly
experimental feature will be called ``filter''\index{norm}{Filter} in the following. 
The calculation of oscillation
probabilities is, in principle, a computation of phase differences. Restricting the maximum admissible size of those phase differences effectively filters
the high frequency component of the oscillation probability. This idea is
implemented according to
\begin{eqnarray}
\label{eq:filter_a}
P_{\alpha\beta}(E)&=&\sum_{ij}
U_{\alpha j} U^*_{\beta j} U^*_{\alpha i} U_{\beta i} 
e^{-i\Phi_{ij}}\times 
e^{ -\Phi_{ij}^2/\sigma_f(E)^2 }\,,
\end{eqnarray}
where $\Phi_{ij}:=\Delta m_{ij}^2 L/2E$ is the usual phase difference and
the last term is a Gau\ss ian filter with width $\sigma_f(E)$. Choosing
$\sigma_f(E):=\sigma_f^0 \cdot E$ ensures that this filter behaves 
approximately as an energy resolution function with constant width 
$\sigma_e=\sqrt{2}/\sigma_f^0$, \ie\
\begin{equation}
\label{eq:filter_b}
\int d\tilde E\quad P(\tilde E) \frac{1}{\sigma_e\,\sqrt{2\pi}}\,
e^{-\frac{(E-\tilde E)^2}{2\sigma^2_e}}\,.
\end{equation}
The relationship between \eqs~\eqref{eq:filter_a} and~\eqref{eq:filter_b}
is not obvious and connected to the properties of $P_{\alpha\beta}$: 
see \Refs~\cite{Kiers:1996zj,Giunti:2003ax}. This feature works \emph{only} 
for vacuum and constant densities and is controlled
by the filer state variable. In addition, $\sigma_e$ is set by the filter value variable:
\index{aedl}{filter state@{\tt \$filter\_state}}
\index{aedl}{filter value@{\tt \$filter\_value}}
\begin{quote}
{\tt
\$filter\_state = 1\\
\$filter\_value = 2.0\\
}
\end{quote}
would switch the filter feature on and set the width to $2.0\,\mathrm{GeV}$.
The setting of {\tt \$filter\_state} is ignored whenever a density profile
with more than one layer is used.

>>>>

\begin{table}[t]
\begin{center}
\begin{tabular}{|cllc|}
\hline
{\tt @builtin} & Description & Parameters & Min. version \\
\hline
1 & Neutrino factory $\mu^+$ decay & {\tt @parent\_energy} [GeV], & 2.0 \\
& & {\tt @stored\_muons} &  \\
2 & Neutrino factory $\mu^-$ decay & {\tt @parent\_energy} [GeV], & 2.0 \\
& & {\tt @stored\_muons} &  \\
3 & Beta beam, $\beta^+$ decay & {\tt @end\_point} [GeV], & 3.0 \\
& & {\tt @stored\_ions}, &  \\
& & {\tt @gamma} & \\
4 & Beta beam, $\beta^-$ decay & {\tt @end\_point} [GeV], & 3.0 \\
& & {\tt @stored\_ions}, &  \\
& & {\tt @gamma} & \\
\hline
\end{tabular}
\end{center}
\mycaption{\label{tab:binfluxes} Built-in fluxes currently supported by \GLOBES . For details on
the beta beams, see \Ref~\cite{Huber:2005jk}. }
\end{table}

@@

\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}\index{norm}{Flux!file}
The software assumes that the given flux file has seven columns and
501 lines\footnote{The requirement of exactly $501$ lines 
has been removed in versions 3.1.7 and higher.} with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by \GLOBES. In general, it is advisable to provide the flux
between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}),
since this is the energy range considered in the simulation. However,
if part of this interval is omitted in the flux file, zero will be
used there. If some neutrino flavors are not used in the simulation,
the corresponding columns in the flux file have to be filled nevertheless,
\eg\ with zeros.

@@

Cross sections in \GLOBES\ are given as total cross section divided by energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines\footnote{The requirement of exactly $1001$ lines 
has been removed in versions 3.1.7 and higher.}
of the form
\index{norm}{Cross section!file}
\index{aedl}{cross@{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Cross sections for unused neutrino flavors have to be filled with zeros, and
can not just be omitted.

@@

\item[Bin level] This level is determined by the experiment and its analysis. 
Note that energy
bin sizes much smaller than the energy resolution will not improve the results. The energy bin range and the number of energy bins always have to be specified. For the case of large values of the integrand in \eq~(\ref{eq:simple_int}) at the energy range limits, it is recommended to exceed the analysis energy window by about three times the energy calibration error in order 
to avoid cutoff effects. In addition, note that an energy resolution larger than about the
bin range will distribute events out of this range. Therefore, the normalization
will be affected.

@@

The key idea is to use a ``flat'' model, \ie\ the integrand 
of~\eq~\ref{eq:simple_int}  is well approximated by being piecewise
constant in each sampling step. This is a good approximation as long as
\begin{itemize}
\item
 No details are lost, \ie\ the spacing of sampling points 
is smaller than the energy resolution.
\item
The edges are treated correctly.
\item
 The neutrino oscillations are slow on a scale of the sampling point distance.
\end{itemize}
In this case, \eq~(\ref{eq:simple_int}) is reduced to
\begin{equation}
\label{eq:algo_one}
n_i^c=N/L^2 \, \sum_{j=1}^N \,  \Phi^c(E_j)\,
P^c(E_j)\,
\sigma^c(E_j)\,
K_i^c(E_j) \, \Delta E_j \,.
\end{equation}
The advantages of this algorithm are obvious: All factors
independent of the oscillation parameters have to be only evaluated once
at values of $E$ which are known in advance, which means that they can be put 
into a look-up table. In addition, the probability
has to be only evaluated at previously known values of the energy, which
makes it possible to compute the transition amplitudes for all channels
simultaneously. One assumption is that all involved factors are piece-wise
constant, \ie, they hardly change within each bin. This 
assumption seems to be very restrictive, which is however not quite correct.
First of all, if one analyzes simulated data (which are simulated with the same algorithm), the errors
will cancel between the simulated and fitted data. Second, and more important,
this algorithm is just a very basic integration routine\footnote{It is planned 
for the future to implement something like a Gau\ss-Kronrod scheme as an
alternative here.} and converges to the true result for decreasing step size.
Thus if the number of sampling points is large enough this algorithm is very
accurate. Bin-based energy smearing is selected by
%
\index{aedl}{energy@{\tt energy}!type@{\tt "@type}}
%
\begin{quote}
{\tt \tb @type = 1}
\end{quote}
within the {\tt \#energy} environment.
The computation of the bin kernel $K_i^c$ is performed
by \GLOBES. Thus, it requires that the number of bins
{\tt \$bins} and the minimum energy {\tt \$emin} and maximum energy {\tt \$emax} are given in case of equidistant bins.
%
\index{norm}{Energy!resolution}
\index{norm}{Energy!resolution function}
%
As far as the parameterization for the energy resolution function 
$R^c(E,E')$ in \eq~(\ref{eq:kernel}) is concerned, the algorithm uses
a Gau\ss ian
\begin{equation}
R^c(E,E')=\frac{1}{\sigma(E)\,\sqrt{2\pi}}\,e^{-\frac{(E-E')^2}{2\sigma^2(E)}} \, .
\label{eq:respfun}
\end{equation} 
%
\index{aedl}{energy@{\tt energy}!sigma function@{\tt "@sigma\_function}}
%
There are several energy resolution functions available, where by default
{\tt \#standard} is used:
\begin{quote}
{\tt \tb @sigma\_function = \#standard} 
\end{quote}
%
\index{aedl}{energy@{\tt energy}!standard@{\tt \#standard}}
%
The energy resolution function {\tt \#standard} is defined by
\begin{equation}
\label{eq:sigma_e}
\sigma(E)=\alpha\cdot E + \beta \cdot \sqrt{E} +\gamma\, ,
\end{equation}
where the parameters $\alpha, \beta$ and $\gamma$ are provided by the user:
\begin{quote}
{\tt \tb @sigma\_e = \{0.15, 0.0, 0.0\}}
\end{quote}
Currently, another possible choice for {\tt @sigma\_function} is {\tt \#inverse\_beta},
%
\index{aedl}{energy@{\tt energy}!inverse beta@{\tt \#inverse\_beta}}
%
which only uses the parameter $\alpha$. It is defined by
\begin{equation}
\sigma(E)= \left\{\begin{array}{cl}
 \alpha \cdot \sqrt{1000}^{-1}\,\sqrt{x-8\cdot10^{-4}}\,,&\mathrm{for}\,\, 
x>1.8\cdot10^{-3}\\
\alpha\cdot10^{-3} \,,&\mathrm{for}\,\, x \leq 1.8\cdot10^{-3}
\end{array} \right.
\end{equation}
The somewhat complicated form is due to the fact that inverse $\beta$-decay
has a neutrino threshold of $1.8\,\mathrm{MeV}$ and that a neutrino
at threshold already produces $\simeq 1\,\mathrm{MeV}$ visible energy in
the detector (for more details see \eg~\cite{Huber:2003pm}).

@@

In order to ensure that very fast oscillations do not lead to 
aliasing,\index{norm}{Aliasing}
it is possible to impose a low-pass filter already during the calculation
of the probabilities itself. This in the meanwhile frequently used feature will be called ``filter''\index{norm}{Filter} in the following. 
The calculation of oscillation
probabilities is, in principle, a computation of phase differences. Restricting the maximum admissible size of those phase differences effectively filters
the high frequency component of the oscillation probability. This idea is
implemented according to
\begin{eqnarray}
\label{eq:filter_a}
P_{\alpha\beta}(E)&=&\sum_{ij}
U_{\alpha j} U^*_{\beta j} U^*_{\alpha i} U_{\beta i} 
e^{-i\Phi_{ij}}\times 
e^{ - \sigma_e^2 \, \Phi_{ij}^2/(2 E^2) }\,,
\end{eqnarray}
where $\Phi_{ij}:=\Delta m_{ij}^2 L/2E$ is the usual phase difference and
the last term is a Gau\ss ian filter with width $\sigma_e$ [GeV]; for details,
see  theory of neutrino oscillations with wave packets, \eg, 
\Refs~\cite{Kiers:1996zj,Giunti:2003ax}. In fact, it can be shown that this filter
corresponds to \eq~(\ref{eq:respfun}) with a constant function $\sigma(E)=\sigma_e$.
 This feature works 
for vacuum and constant densities \emph{only}, and it is controlled
by the filer state variable. In addition, $\sigma_e$ is set by the filter value variable:
\index{aedl}{filter state@{\tt \$filter\_state}}
\index{aedl}{filter value@{\tt \$filter\_value}}
\begin{quote}
{\tt
\$filter\_state = 1\\
\$filter\_value = 2.0\\
}
\end{quote}
would switch the filter feature on and set the width to $2.0\,\mathrm{GeV}$.
The setting of {\tt \$filter\_state} is ignored whenever a density profile
with more than one layer is used.
